Improving \textsl{Gaia} parallaxes with a data-driven model \\ of the color--magnitude diagram

### 2017-06-02 paper/paper.tex

% # To-do list
% - We need a standard name for the \tgas--\tmass--\PanSTARRS\ Intersection.
% - deal with all DWH or LMA in the text.
% - deal with everything in the google doc.
% - are there any github issues?

@@

We need to correct the photometry used in our method, both for the generation of the prior and the likelihood, for dust. With the emergence of 3D dust maps, it is now possible to estimate dust corrections to stars within the Milky Way. The challenge is getting a proper estimate of the distance to the star before inferring $\varpi_{\true}$, since we need a dust correction prior to doing our inference. Dust corrections are particularly non trivial for stars with poor parallax signal to noise. Stars with poor signal to noise have a likelihood that is consistent with infinite distance and can therefore have severe dust corrections. This is most obvious in the giant stars, which tend to have the lowest signal to noise, and can therefore have dust corrections $> 2$ magnitudes. Although the dust model is probabilistic, as well as our distance inference, to move forward, we break from a fully Bayesian framework, in which we might infer the distance and the dust simultaneously (more about this in the discussion). Instead, we correct for dust using a point estimate from the 3D dust map \citep{green15} sampled at a distance estimated from iteratively inferring and then sampling our posterior pdf over $\varpi_{\true}$. In more detail, to apply a dust correction to the \tmass\ photometry, we first generate our prior using the observed, attenuated photometry. Using this raw prior, we infer more precise parallaxes to all the stars in our sample. We then use this more precise parallax posterior to get a measurement of dust in the 3D dust map for each star. And repeat. This approach minimizes severe distances, and their associated severe dust corrections, which would occur if using the likelihoods alone. In detail, we take the $5\%$ quantile of each distance posterior, so the closest bit of the posterior, and sample the 3D dust map at that distance. We take the $5\%$ quantile, as apposed to the median, for example, to do a light dust correction, because over correcting for dust can shoot the likelihood into an unphysical, different space of the \cmd\ and bias the parallax inference. For each distance, we sample the median of the probabilistic dust model. We apply each light dust correction to our \tmass\ photometry given by Equation \ref{eq:data} where $Q_{\lambda}$ = [0.709, 0.302] for bands [J, K] respectively, taken from table (LMA REF TABLE SFD). We then regenerate our prior using the lightly dust corrected photometry. The dust values we obtain in our final iteration, and which we apply to our photometry, are visualized in galactic coordinates in Figure \ref{fig:dust}. After applying a dust correction, we do not update the uncertainties, but keep them the same as before the dust correction, which is wrong and under representing our uncertainties but this leads, again, to a lighter deconvolution. 
In other words, the features of our inferred \cmd\ are conservatively broad (broader than if we over-estimated the errors which would have lead to excessive deconvolution and unrealistically narrow features). We could add the covariance of the dust to our likelihoods, but this is beyond the scope of this demonstration of concept. We iterate this process 10 times, however the dust values seem to converge after about 2-3 iterations. The iterative process allows for slight variations in distance and corresponding dust to become more consistent with the \cmd.
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{dust.png}
\caption{The converged dust values at the 5\% distance projected on the l,b galactic coordinates. This shows both the footprint of our analysis as well as the dust values we used. Each point represents a single star in our cross matched catalog. The large missing areas are due to the lack of dust values from \panstarrs\ since it only looked above -30 degrees in dec }
\label{fig:dust}
\end{figure}

LMA: Should I quote some statistics about the dust corrections? mostly small? maybe difference before/after applying prior to distance?

@@

Figure~\ref{fig:posteriorCMD} shows the distribution of the posterior parallax inferences derived with our \cmd\ prior.
This is further illustrated in Figure~\ref{fig:delta}, where we show the fractional changes, and highlight the regions of \cmd\ space with the greatest improvement.
Those typically are the regions with the largest photometric and parallax errors, for the faintest objects.
This is a fairly standard result: the deconvolution of uncertainties is stronger in noisier regions of the data, leading to narrower features in the upper part of the \cmd, for example.

@@

\begin{figure}
\centering
  \includegraphics[width=\textwidth]{posterior.png}
\caption{5 example posteriors showing the priors and likelihoods as well}
\label{fig:posterior}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth]{deltaCDF.png}
\caption{Fractional Change in Variance: The cumulative distribution function of the fractional variance change when comparing our posterior to the \gaia\ catalogue.}
\label{fig:deltaCDF}
\end{figure}

\subsection{Distances to M67}
As a test of the accuracy of these photometric parallax inferences, we inferred the parallaxes to stars in the open cluster M67.It is an open cluster in the constellation Cancer and is estimated to be 3-5 billion years old and 800-900 pc away. Previous work (LMA cite CBJ) has estimated that bad things happen if you try to infer a distance as the inverse of the observed (noisy) parallax when the signal to noise of the parallax measurement drops below 5. With \gaia\ parallaxes having a minimum uncertainty of about 0.3 mas, the SN reaches 5 for the most certain parallax measurements at about 700 pc, right where this cluster lies. So it is a great test case for comparing parallax inferences with or without a prior. We made a window of 1 square degree on the sky with the cluster M67 lying in the middle. The cluster is visibly obvious as a clustering of stars in a visualization of the \gaia\ data on the 2D sky. With our converged dust values, and its associated prior, we calculate the posterior over $\varpi_{\true}$ for each star within this window on the sky around M67. Figure \ref{fig:m67} shows the comparison of the likelihoods and posteriors. The vertical lines bracket the previously measured parallaxes of, or distances to, the cluster. The likelihoods of the observed parallaxes, shown in the top row, are broad and the cluster is not obvious at all. The posterior pdf, shown in the bottom row, are more sharply peaked at the previously determined distance to the cluster, the vertical lines, for far more stars than the likelihoods. Our prior is not only making parallaxes more precise but also (possibly) increasing the accuracy.
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{distancesM67.png}
\caption{Proof of concept: parallax inference to the open cluster M67}
\label{fig:m67}
\end{figure}

@@

\item[stationarity] We have assumed that (locally) the morphology of

@@

  Beyond this, the choice of the Gaussian form itself for the mixture is
  questionable. The Gaussian has great properties, that we capitalize on
  both at the prior-generation stage (with \xd), and at the photometric-parallax
  stage (capitalizing on Gaussian product rules). We could move to other
  forms, but they would be extremely expensive, computationally.
\item[no physics] Although it is presented as an advantage of our
  method that we never make any use of physical models, it is also
  brings some significant disadvantages: For example, the \cmd\ contains
  features that are obvious products of stellar evolution, but doesn't in
  itself inform the theory or models of stellar evolution, because our
  mixture-of-Gaussian \cmd\ model is divorced from any theoretical ideas
  about stars. For another, stellar evolution models are in fact very
  predictive and successful; our model does not capitalize in any way on
  those successes. A next-generation model might be designed to be data-driven
  but nonetheless capitalize on physics-driven successes. One framework
  for that would be to fit not the \cmd\ but rather deviations of the \cmd\ away
  from a good physics-based model prediction.
\end{description}

>>>>

@@

We need to correct the photometry used in our method, both for the
generation of the prior and the likelihood, for dust. With the
emergence of 3D dust maps, it is now possible to estimate dust
corrections to stars within the Milky Way. The challenge is getting a
proper estimate of the distance to the star before inferring
$\varpi_{\true}$, since we need a dust correction prior to doing our
inference. Dust corrections are particularly non trivial for stars
with poor parallax signal to noise. Stars with poor signal to noise
have a likelihood that is consistent with infinite distance and can
therefore have severe dust corrections. This is most obvious in the
giant stars, which tend to have the lowest signal to noise, and can
therefore have dust corrections $> 2$ magnitudes. Although the dust
model is probabilistic, as well as our distance inference, to move
forward, we break from a fully Bayesian framework, in which we might
infer the distance and the dust simultaneously (more about this in the
discussion). Instead, we correct for dust using a point estimate from
the 3D dust map \citep{green15} sampled at a distance estimated from
iteratively inferring and then sampling our posterior pdf over
$\varpi_{\true}$. In more detail, to apply a dust correction to the
\tmass\ photometry, we first generate our prior using the observed,
attenuated photometry. Using this raw prior, we infer more precise
parallaxes to all the stars in our sample. We then use this more
precise parallax posterior to get a measurement of dust in the 3D dust
map for each star. And repeat. This approach minimizes severe
distances, and their associated severe dust corrections, which would
occur if using the likelihoods alone. In detail, we take the $5\%$
quantile of each distance posterior, so the closest bit of the
posterior, and sample the 3D dust map at that distance. We take the
$5\%$ quantile, as apposed to the median, for example, to do a light
dust correction, because over correcting for dust can shoot the
likelihood into an unphysical, different space of the \cmd\ and bias
the parallax inference. For each distance, we sample the median of the
probabilistic dust model. We apply each light dust correction to our
\tmass\ photometry given by Equation \ref{eq:data} where $Q_{\lambda}$
= [0.709, 0.302] for bands [J, K] respectively, taken from Table 6 of
cite{schlafly11}. We then regenerate our prior using the lightly dust
corrected photometry. The dust values we obtain in our final
iteration, and which we apply to our photometry, are visualized in
galactic coordinates in Figure \ref{fig:dust}. After applying a dust
correction, we do not update the uncertainties, but keep them the same
as before the dust correction, which is wrong and under representing
our uncertainties but this leads, again, to a lighter
deconvolution. In other words, the features of our inferred
  \cmd\ are conservatively broad (broader than if we over-estimated
  the errors which would have lead to excessive deconvolution and
  unrealistically narrow features). We could add the covariance of
the dust to our likelihoods, but this is beyond the scope of this
demonstration of concept. We iterate this process 10 times, however
the dust values seem to converge after about 2-3 iterations. The
iterative process allows for slight variations in distance and
corresponding dust to become more consistent with the \cmd.
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{dust.png}
\caption{The converged dust values at the 5\% distance projected on the l,b galactic coordinates. This shows both the footprint of our analysis as well as the dust values we used. Each point represents a single star in our cross matched catalog. The large missing areas are due to the lack of dust values from \panstarrs\ since it only looked above -30 degrees in dec }
\label{fig:dust}
\end{figure}

Should I quote some statistics about the dust corrections? mostly small? maybe difference before/after applying prior to distance?

@@

Figure~\ref{fig:posteriorCMD} shows the distribution of the posterior parallax inferences derived with our \cmd\ prior.
This is further illustrated in Figure~\ref{fig:delta}, where we show the fractional changes, and highlight the regions of \cmd\ space with the greatest improvement.
Those typically are the regions with the largest photometric and parallax errors, for the faintest objects.
This is a fairly standard result: the deconvolution of uncertainties is stronger in noisier regions of the data, leading to narrower features in the upper part of the \cmd, for example.
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{posterior.png}
\caption{5 example posteriors showing the priors and likelihoods as well}
\label{fig:posterior}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\textwidth]{deltaCDF.png}
\caption{Fractional Change in Variance: The cumulative distribution function of the natural log of the fractional variance change when comparing our posterior to the \gaia\ catalogue. Negative numbers represent stars which have a posterior that is narrower than the \gaia\ catalogue, which is true for a large fraction of the stars. We've drawn two dashed lines. One represents the median star $(y=0.5)$ which has a fractional variation of $x=-0.3=ln(1/1.2^2)$, corresponding to an increase in precision by a factor of 1.2. Another represents the fraction of stars $(y=0.14)$ which have a fractional variance greater than $x=-1.4=ln(1/2^2)$, corresponding to an increase in precision by a factor of 2.}
\label{fig:deltaCDF}
\end{figure}

@@

\subsection{Distances to M67}
As a test of the accuracy of these photometric parallax inferences, we inferred the parallaxes to stars in the open cluster M67.
It is an open cluster in the constellation Cancer and is estimated to be 3-5 billion years old and 800-900 pc away.
Previous work \citep{cbj15} has estimated that bad things happen if you try to infer a distance as the inverse of the observed (noisy) parallax when the signal to noise of the parallax measurement drops below 5.
With \gaia\ parallaxes having a minimum uncertainty of about 0.3 mas, the SN reaches 5 for the most certain parallax measurements at about 700 pc, right where this cluster lies.
So it is a great test case for comparing parallax inferences with or without a prior.
We made a window of 1 square degree on the sky with the cluster M67 lying in the middle.
The cluster is visibly obvious as a clustering of stars in a visualization of the \gaia\ data on the 2D sky.
With our converged dust values, and its associated prior, we calculate the posterior over $\varpi_{\true}$ for each star within this window on the sky around M67.
Figure \ref{fig:m67} shows the comparison of the likelihoods and posteriors.
The vertical lines bracket the previously measured parallaxes of, or distances to, the cluster.
The likelihoods of the observed parallaxes, shown in the top row, are broad and the cluster is not obvious at all.
The posterior pdf, shown in the bottom row, are more sharply peaked at the previously determined distance to the cluster, the vertical lines, for far more stars than the likelihoods.
Our prior is not only making parallaxes more precise but also (possibly) increasing the accuracy.
\begin{figure}
\centering
  \includegraphics[width=\textwidth]{distancesM67.png}
\caption{Proof of concept: parallax inference to the open cluster M67}
\label{fig:m67}
\end{figure}

@@

\item[stationarity]
  We have assumed that (locally) the morphology of

@@

  Beyond this, the choice of the Gaussian form itself for the mixture is
  questionable. The Gaussian has great properties, that we capitalize on
  both at the prior-generation stage (with \xd), and at the photometric-parallax
  stage (capitalizing on Gaussian product rules). We could move to other
  forms, but they would be extremely expensive, computationally.
\item[no physics] Although it is presented as an advantage of our
  method that we never make any use of physical models, it is also
  brings some significant disadvantages: For example, the 
\end{description}
